---
title: "Rpart_examples"
author: "Raymaekers, J. and Rousseeuw, P.J."
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rpart_examples}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
 fig.width  = 5 ,
 fig.height = 3.5,
 fig.align  = 'center'
)
oldpar <- list(mar = par()$mar, mfrow = par()$mfrow)
```

# Introduction

This file illustrates the silhouette plot and class map for visualizing classification results from classification trees fit by rpart. The discussion of the results of this vignette is in section 4 of Raymaekers and Rousseeuw (2021) (for the training data), and in section A.4 of the Supplementary Material (for the test data).

```{r}
library(rpart)
library(classmap)
```


# Titanic training data 

We first load and inspect the data.

```{r}
data("data_titanic")

traindata = data_titanic[which(data_titanic$dataType == "train"), -13]

dim(traindata) 
colnames(traindata)
# SibSp: number of siblings/spouses aboard
# Parch: number of parents/children aboard

str(traindata)
table(traindata$y)
```

Now we want to predict the response y by CART (rpart).
We set the seed as rpart is not deterministic.
We also inspect the resulting classification tree.

```{r}
set.seed(123) 
rpart.out = rpart(y ~ Pclass + Sex + SibSp + 
                    Parch + Fare + Embarked,
                  data=traindata, method='class', model=T)

## plot the tree:
# pdf(file = "titanic_train_tree.pdf", width=6, height=6)
rpart.plot::rpart.plot(rpart.out)
# dev.off()
```

We now inspect the variable importance, which will be used to calculate
the variable weights in the farness computation. 

```{r}
rpart.out$variable.importance
```

Now we declare the types of the variables. This is used in the calculation of the Daisy distance, which in turn is needed for the farness computation. There are 5 nominal columns and one ordinal. The variables not listed here are interval-scaled by default.
```{r}
mytype = list(nominal=c("Name","Sex","Ticket","Cabin","Embarked"), 
              ordratio=c("Pclass"))
```

Now we prepare for visualization and inspect the elements of the output.

```{r}
x_train = traindata[,-12]
y_train = traindata[, 12]
vcrtrain = vcr.rpart.train(x_train, y_train, rpart.out, mytype)
names(vcrtrain)

vcrtrain$predint[1:10] # prediction as integer
vcrtrain$pred[1:10]    # prediction as label
vcrtrain$altint[1:10]  # alternative label as integer
vcrtrain$altlab[1:10]  # alternative label

# Probability of Alternative Class (PAC) of each object:
vcrtrain$PAC[1:3] 
#
summary(vcrtrain$PAC)

# f(i,g) is the distance from case i to class g:
vcrtrain$fig[1:3,] # for the first 3 objects:

# The farness of an object i is the f(i,g) to its own class: 
vcrtrain$farness[1:3]
#
summary(vcrtrain$farness)

# The "overall farness" of an object is defined as the 
# lowest f(i,g) it has to any class g (including its own):
summary(vcrtrain$ofarness)

sum(vcrtrain$ofarness > 0.99, na.rm = T) 
# No farness is considered outlying in these data.

confmat.vcr(vcrtrain) 

cols = c("firebrick", "blue")
```

We first make the stacked plot, visualizing the confusion matrix:

```{r}
stackedplot(vcrtrain, classCols = cols)
```

Now we consider the silhouette plot

```{r}
silplot(vcrtrain, classCols = cols)
# silplot.out <- silplot(vcrtrain, classCols = cols)
# ggsave("titanic_train_silhouettes.pdf", silplot.out,
#        width = 5, height = 5)
```


Now we construct the quasi residual plot of PAC vs. age for males only.
The age variable is not in the model, but including it
did not improve the classification. We see that the very young male passengers are often misclassified.
```{r}
hist(x_train$Age)

# Quasi residual plot versus age, for males only:
xm = x_train$Age[which(x_train$Sex == "male")]
ym = vcrtrain$PAC[which(x_train$Sex == "male")]
#
# pdf("titanic_qrp_versus_age_males.pdf", width=4.8, height=4.8)
par(mar = c(3.5, 2, 2, 0))
par(pty="s") # makes the axes of the plot equally long
plot(xm, ym, type= "n", ann = FALSE, axes = FALSE)
# makes an empty plot
title(main = "quasi residual plot for males", line = 1, 
      cex.main = 1.2)
title(ylab = "P[alternative class]", line = 2.3, 
      cex.lab = 1)
title(xlab = "Age (years)", line = 2.3, cex.lab = 1)
par(new=TRUE) # overlay next plot on this one
plot(xm, ym, xlab = "", ylab = "")
lom = loess(ym ~ xm)
lines(0:82, predict(lom, 0:82), col="red", lwd = 2)
text(x=17,y=0.56,"loess curve",col="red", cex = 1)
# dev.off()

ind = which(!is.na(xm))
cor.test(xm[ind], ym[ind], method="spearman")
# p-value is 0.0961 so not significantly different from zero.
# However, loess shows the local effect at very young age.
```

We can do the same for females, where the effect of age is clearly much smaller:
```{r}
xf = x_train$Age[which(x_train$Sex == "female")]
yf = vcrtrain$PAC[which(x_train$Sex == "female")]
plot(xf,yf)
lof = loess(yf ~ xf)
lines(0:82, predict(lof, 0:82), col="red")

```


Now we construct the class maps. First of the casualties.

```{r}
classmap(vcrtrain, "casualty", classCols = cols)
# classmap(vcrtrain, "casualty", classCols = cols, identify = T)

# blue points top right: cases "a" and "b" in the paper
x_train[which(y_train == "casualty")[119], ]
# Woman in 1st class, should have survived.
#
x_train[which(y_train == "casualty")[192], ]

# Similar, but child.

# red point most to the right: case "c" in the paper
x_train[which(y_train == "casualty")[268], ]

```

Now the class map of the survivors.

```{r}
classmap(vcrtrain, "survived", classCols = cols)
# classmap(vcrtrain, "survived", classCols = cols, identify = T)

# red point with highest farness among highest PAC: case "d" in the paper
x_train[which(y_train == "survived")[c(14)], ] 

# near-coinciding points with highest farness: cases "e" and "f" in the paper
#
x_train[which(y_train == "survived")[265], ]

x_train[which(y_train == "survived")[287], ] 

# man --> predicted as not survived. also paid the highest 
# fare in the data and has same ticket number as passenger 
# 265. Also embarked as same place. these may be related?
# according to following links they were colleagues working 
# for the Cardeza family:
# https://www.encyclopedia-titanica.org/titanic-survivor/thomas-cardeza.html
# https://www.encyclopedia-titanica.org/titanic-survivor/gustave-lesueur.html

# blue point bottom right: case "g" in the paper
x_train[which(y_train == "survived")[90], ]

# # woman  + first class so predicted in survivors.
# # paid highest fare in whole dataset 
```

# Titanic test data 
In addition to the training data, we consider the titanic test data.
First we load the data and inspect.


```{r}
testdata = data_titanic[which(data_titanic$dataType == "test"), -13]

dim(testdata)
x_test = testdata[, -12]
y_test = testdata[, 12]
table(y_test)
```

Now we prepare for visualization:

```{r}
vcrtest = vcr.rpart.newdata(x_test, y_test, vcrtrain)

confmat.vcr(vcrtest)
cols = c("firebrick", "blue")
``` 

Now we can visualize, starting with the stacked plot:
```{r}
stackedplot(vcrtest, classCols=cols)
```

Now we construct the silhouette plot:

```{r}
silplot(vcrtest, classCols = cols)
```

Finally, we make the class maps. First of the casualties, in which the misclassifications are all female (since all males 
are predicted as casualty by this model).

```{r}
classmap(vcrtest, "casualty", classCols = cols) 
```

Now the class map of survivors:

```{r}
classmap(vcrtest, "survived", classCols = cols) 
```

***
